---
title: "Lecture 6 Notes: Iterations, complexity, and syntactic sugar"
author: "Jillian Allison"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(palmerpenguins)
library(tidyverse)
```

### Referring to columns in a data frame

```{r results='hide'}
# reference a column using 
penguins$species 

```

### For loops in mtcars data frame

```{r}
# finding second column of mtcars --- number of cylinders 
mtcars_df2 <- mtcars[2] # outputs 2 columns... car type and cylinder column -- class = data frame
mtcars_vector2_1 <- mtcars[,2] # outputs a vector which we can do calculations on -- class = vector
mtcarsvector2_2 <- mtcars[[2]] # also outputs a vector which we can do calculations on -- class = vector


# Using this, we can now refer to different columns in a data frame and loop through them! 

# Lets iteratively calculate the mean of each column in mtcars 
mean_mtcars <- vector(mode = "numeric", length = ncol(mtcars))

for (i in seq_along(mtcars)) {        
                    # i starts at 1 and makes a value for each column of mtcars
  mean_val <- mean(mtcars[[i]], na.rm = TRUE) 
                    # storing the mean of the i'th column of mtcars in intermediate object mean_val 
  mean_mtcars[i] <- mean_val
                    # storing mean value for the i'th column from mean_val object and populating i'th
                    # position in mean_mtcars vector 
}
```

What if we don't want to iterate across all columns? Let's try only iterating across numeric columns in palmerpenguins!

```{r}
# create a vector called penguin_med
penguin_med <- vector(mode = "numeric", length = ncol(penguins)) 
# except we'd want only the columns in the penguins data frame that are numeric! which would take code that would distract me from class if i tried to write it. so sorry future jill

# finding median for each numeric column in penguins data frame 

for(i in seq_along(penguins)) { 
  if(is.numeric(penguins[[i]])) # conditional statement inside of if()
  penguin_med <- median(penguins[[i]], na.rm = TRUE)
  print(penguin_med)
  }
```

### Functional programming // syntactic sugar!

use functions to write your program .... don't reinvent the wheel

3 options for functions:

**apply()**:a family of functions in base R, helpful for iterating across rows or columns

\>\>apply a function across rows or columns of a matrix (or data frame)

*apply function arguments*:

apply(data, (1 is row, 2 is column), function to be applied)

```{r}
apply(mtcars, 2, mean) # applying mean function to columns in mtcars data frame 
```

**tapply():** part of apply(), apply a function by group (factor level)

```{r}
# tapply(penguins_size$flipper_length_mm, penguins_size$species, mean, na.rm = TRUE)
```

**dplyr::across()** : a function in dplyr to apply a function across specified columns

\>\> use within mutate() or summarize() to apply functions to a selection of columns

mutate() adds or changes columns.

```{r results='hide'}
penguins %>% 
  mutate(across(c(species, island), toupper))
```

**{purrr}**: an R package with "the family of map() function swhich allow you to replace many for loops with code that is both more succint and easier to read"

## Afternoon active coding sesh 

#### group_by() %\>% summarize() -- amazing summary data!

In penguins, calculate mean bill length by species.

Starting from penguins. then under the hood, tell R to recognize different groups within species. then we want to calculate the mean and stdev bill length for each group, and then we want it to return in a nice table.

```{r}
penguins_BL_summary <- penguins %>% 
  group_by(species, island) %>%  
        # under the hood, now R knows that whatever we do next, we're assuming that we want          to apply it by group species 
  summarize(mean_bill_length = mean(bill_length_mm, na.rm = TRUE), 
            sd_bill_length = sd(bill_length_mm, na.rm = TRUE), 
            n = n()) # adds sample size column-- counts number of samples in each row 
        # apply given function by group and returns it in a nice table 

  
unique(penguins$species) # gives us unique species values in penguins data frame 
```

#### More group_by() %\>% summarize() examples

```{r}
starwars <- starwars
# Find the mean and sd of character height in "starwars" grouped by homeworld 
character_height <- starwars %>% 
  group_by(homeworld) %>% 
  summarize(mean_height = mean(height, na.rm = TRUE),
            sd_height = sd(height, na.rm = TRUE)) 
  
# Find the maximum & minimum flipper length, and sample size for each species of penguins 
penguins <- penguins
flipper_length <- penguins %>% 
  group_by(species, sex) %>%
  summarize(maximum_flipper_length = max(flipper_length_mm, na.rm=TRUE),
            minimum_flipper_length = min(flipper_length_mm, na.rm= TRUE),
            n = n())

penguins %>% 
  group_by(species) %>%
  summarize(across(where(is.numeric), mean, na.rm = TRUE))
```

### back to for loops! for loop with cumulative calculations

```{r}
# cumulative sums-- adding next value to sum from previous value ...? 
tree_height <- c(1, 2, 6, 10, 14, 20)

for (i in seq_along(tree_height)) {
  val = tree_height[i] + tree_height[i + 1]
  print(val)
}


```

```{r}
# converting human years to animal years
species <- c("dog","elephant","goat","dog","dog","elephant") 

age_human <- c(3, 8, 4, 6, 12, 18)

animal_ages <- vector(mode = "numeric", length = length(species))

for (i in seq_along(species)) {
  if (species[i] == "dog") {
    animal_age <- age_human[i] * 7 
  }
  else if(species[i] == "elephant"){
    animal_age <- age_human[i] * 0.88 
  }
  else if(species[i] == "goat"){
    animal_age <- age_human[i] * 4.7
  }
  animal_ages[i] <- animal_age
}
```

### for loops across data frames

```{r}
mean_mtcars_floop <- vector(mode = "numeric", length = ncol(mtcars))
for (i in seq_along(mtcars)) {
  mean_val <- mean(mtcars[[i]], na.rm = TRUE)
  mean_mtcars_floop[i] <- mean_val
}
mean_mtcars_floop
```

### Functional programming examples-- syntactic sugar 

apply, across, map_df

#### Apply

```{r}
apply(X = mtcars, MARGIN = 2, FUN = mean)
```

#### {Purr} 

```{r}
# list
map(.x = mtcars, .f = mean) # .x is data argument, .f is function argument
# data frame 
map_df(.x = mtcars, .f = mean)
```
