---
title: "Lecture 8: Functions continued - planning, building,  adding complexity, useful
  warnings and error messages"
author: "Jillian Allison"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

### DIY nested for loop

```{r}
length <- c(2, 4, 7, 9)
width <- c(3, 1, 6, 4)

area_matr <- matrix(nrow = length(length), ncol = length(width))

for (i in seq_along(length)){
  for (j in seq_along(width)) {
    area <- length[i] * width[j]
    area_matr[i,j] <- area
  }
}

guava_price <- function(x, guava) {
  5 * x - guava
} 
# SHORTCUT COMMAND OPTION X --- WRITE OUT INSIDE OF FUNCTION AND IT CREATES STRUCTURE WITH INPUTS RLY FAST!! HOW EXCITING IS THAT????
```

LOGISTIC GROWTH

Creating logistic growth function

```{r}
logistic_growth <- function(K, N0, r, time) {
  K / (1 + ((K - N0)/N0) * (exp(-r * time)))
}
```

Try function through time_vec

```{r}
time_vec <- seq(from = 0, to = 35, by = 0.1)

pop_35 <- logistic_growth(N0 = 2500, K = 4000, r = 0.25, time = time_vec)

pop_time35_df <- data.frame(time_vec, pop_35)
```

ggplot the population over time!!

```{r}
pop_time35_gg <- ggplot(data = pop_time35_df, aes(x = time_vec, y = pop_35)) +
  geom_point()

pop_time35_gg
```

doing the same thing with a for loop instead

```{r}
pop_35_vec <- vector(mode = "numeric", length = length(time_vec))

for (i in seq_along(time_vec)) { 
  population <- logistic_growth(N0 = 100, K = 6000, r = 0.27, time = time_vec[i])
  pop_35_vec[i] <- population
  }
```

Now extend this to iterate over multiple growth rates.

```{r}
r_seq <- seq(from = 0.2, to = 0.4, by = 0.01)
```

making our data frame using a nested for loop and function!

```{r}
out_matrix <- matrix(nrow = length(time_vec), ncol = length(r_seq)) 

for(i in seq_along(r_seq)){
  for(j in seq_along(time_vec)){
    pop <- logistic_growth(N0 = 100, K = 6000, r = r_seq[i], time = time_vec[j])
    out_matrix[j,i] <- pop
  }
}

# now some wrangling 
logistic_df <- data.frame(out_matrix, time = time_vec)

colnames(logistic_df) <- c(paste0("gr_", r_seq), "time")

logistic_df_long <- logistic_df %>%
  pivot_longer(cols = -time,
               values_to = "population",
               names_to = "growth_rate")

ggplot(data = logistic_df_long, aes(x = time, y = population)) + 
  geom_line(aes(color = growth_rate)) 
```

### A function with warnings and an error message

use `warning` to add a warning

use `stop` to add an error

P = 0.3(rho) * A * r^2 * v^3 

rho = air density
a = area covered by turbine
v = wind velocity

```{r}

calc_windpower <- function(rho, radius, windspeed) {
  if (windspeed > 130) {
    warning("Whoa, that wind is really fast. Are you sure?")
  } 
  if (rho > 1.225){ 
    warning("That air density is suspicious.")
  }
  if (radius < 0){ 
    stop("Rotor radius must be a positive value in meters")
  }
  0.3*rho*pi*(radius^2)*(windspeed^3)
}

calc_windpower(rho = 1.225, radius = 50, windspeed = 40)

```
