---
title: 'Lecture 11 & 12: tidy data and {tidyr}, {dplyr}'
author: "Jillian Allison"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

##### Warm up 

```{r}
animals <- c("brown bear","red bird","yellow duck","blue horse") 

for (i in seq(from = 1, to = (length(animals) - 1))){
  print(paste(animals[i], animals[i], "what do you see? I see a", animals[i + 1], "looking at me!"))
}
```

## Lecture 11
### Tidy data! 
Tidy data is a predictable way of organizing data where : 

1) each variable is a column. variable = characteristic being measured or described with data. 

2) each observation is a row. observation = a single data point for one or more variables. 

3) each cell contains a single value. value = recorded measure, count, or description of a variable.

it's not R specific! it's just a way to organize data that will make life easier for people working with data. that's it! 

Benefits: 

1) makes collaboration easier 

2) prevents people from "reinventing the wheel"

3) reuse: reuse similar wrangling and analyses across different/updated datasets 

4) generalizability: tools built for one tidy dataset are largely generalizable ^^ similar to 3


### {tidyr} 

- pivot_wider() and pivot_longer() 

example:
```{r}
# PIVOT LONGER
# tidyr::pivot_longer(week_1:week_3, 
                   # names_to = week,
                   # values_to = miles)
#lengthening the week 1 through 3 columns, changing names of the columns to being values in a column called 'week', and putting those values within those previous columns into a column called miles

# PIVOT WIDER
# pivot_wider(names_from = variable column,
#             values_from = value column) 
```

- separate() 

separate elements in a single column into multiple columns, using a separator you specify 

```{r}
# separate(col = name,
         # into = c("first_name","last_name"),
         # sep = " ")
```


## Lecture 12 

### Subsetting rows based on your conditions with filter() 

Example 1: filter to only keep exact matches 
```{r}
ex_1 <- starwars %>%
  filter(height > 180)
```

Example 2: Exact match
```{r}
ex_2 <- starwars %>% 
  filter(eye_color == "blue")
```

Example 3: make a subset that only contains rows where the homeworld is Naboo
```{r}
ex_3 <- starwars %>% 
  filter(homeworld == "Naboo")
```

Examples 4 and 5: Filtering to satisfy multiple conditions 

```{r}
ex_4 <- starwars %>% 
  filter(homeworld == "Tatooine" & height > 180)
```

brown hair color & species is human
```{r}
ex_5 <- starwars %>% 
  filter(hair_color == "brown" & species == "Human")
```

Example 6: Filtering OR statements 
```{r}
ex_6 <- starwars %>% 
  filter(height > 180 | eye_color == "yellow")
```

homeworld == endor or species is droid
```{r}
ex_7 <- starwars %>% 
  filter(homeworld == "Endor" | species == "Droid")
```

eye color is blue or red or brown 
```{r}
ex_8 <- starwars %>% 
  filter(eye_color %in% c("blue","brown","red")) 

## DONT USE == HERE! IT GOES THROUGH THE COLUMN ASKING FOR EXACT MATCHES FOR 1st row = [1] 2nd row = [2] 3rd = [3] 4th = [1] ... and so on. so we like %in% !
```

Example 11 bc I got off at some point 
endor or tatooine or naboo AND 
```{r}
ex_11 <- starwars %>% 
  filter(homeworld %in% c("Endor","Tatooine","Naboo"),
         height > 150)
```

Examples with negate operator 

keep all observations where speces =/= human
```{r}
ex_12 <- starwars %>% 
  filter(species != "Human")

ex_13 <- starwars %>% 
  filter(!species %in% c("Human", "Ewok", "Wookie"))
```

### Subsetting columns with select() --- SEE END OF INTERACTIVE NOTES WEEK2DAY1 !!! 


