---
title: 'Lecture 10: finding, exploring, and using packages'
author: "Jillian Allison"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```


### Finding new packages

-   CRAN task views can be used to find the packages we'd use for data exploration!
-   but really google and twitter

### Exploring and vetting packages

-   there exists a useful checklist for choosing R packages
-   cran projects are generally really good

1)  maturity
2)  active development
3)  recently updated
4)  documentation available (THEY BETTER HAVE IT! NECESSITY)
5)  used/published in similar projects
6)  license
7)  semantics intuitive
8)  functions that get the job done
9)  arguments to support your needs
10) dependencies reported / reasonable

### Citing packages 
-   we love citing other people's stuff! cite packages and openly available resources
-   can add a citation for your own package in roxygen template! 

```{r}
citation("palmerpenguins")
```

##### some examples that are relevant to the activity 

Coercing column classes 
```{r}
penguins_year_chr <- penguins %>% 
  mutate(year = as.character(year))

penguins_year_num <- penguins%>%
  mutate(year = as.numeric(year))

penguins_ad_ch <- penguins_year_num %>% 
  filter(species %in% c("Adelie","Chinstrap"))
```

