---
title: 'Lecture 7: Introduction to functions'
author: "Jillian Allison"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # for tribble function

```
### Coding warm up-- loop! 
```{r}
quarter_splits <- c(1.0, 1.1, 1.2, 1.1, 1.4, 1.5, 1.6, 1.4)
for (i in seq_along(quarter_splits)){
  consec <- quarter_splits[i] + quarter_splits[i + 1]
}
```

### Nested for loop examples 
```{r}
# COMPLETE EVERYTHING IN INNER FOR LOOP BEFORE MOVING ON TO NEXT I VALUE FOR OUTER LOOP
# Example one: prints temp 1 through 5, ph 1 through 5, salinity 1 through 5. 
file_prefix <- c("temp", "ph", "salinity")
file_suffix <- c(1, 2, 3, 4, 5)
for (i in seq_along(file_prefix)) {
  for (j in seq_along(file_suffix)) {
print(paste0(file_prefix[i], "_", file_suffix[j]))
  }
}

# Example two: 
odds <- c(1, 3, 5)
evens <- c(2, 4, 6, 8)
for (i in seq_along(odds)) {
  for (j in seq_along(evens)) {
print(odds[i] * evens[j])
  }
}
```

## Functions
*You should consider writing a function whenever you've copied and pasted a block of code more than twice (you now have 3 copies of the same code!)* 

##### Benefits of writing functions
* give a function an evocative name that makes your code easier to understand
a. probably use verbs to name functions, unless its a very common noun like "mean" 
* as requirements change, you only need to update code in one place, instead of many. 
* you eliminate the chance of making incidental mistakes when you copy and paste (ie changing variable names) 

##### Anatomy of a function 
function_name <- function(arguments, arguments){
body and outputs 
}

then to call / run the function... 
function(argument = "YAY") 

### Example 1: Make a function that doubles user input 
```{r}
double_this <- function(my_value){
  print(2 * my_value)
}

double_this(my_value = 1234)

double_this(c(1, 2, 10, 100, 175))

```
 
### Example 2: Write a function that asks for a user's name and age and returns an exclamation like "I'm NAME and I am AGE years old."

```{r}
say_hi <- function(name, age){
  paste("Good morning! I'm", name, "and I am", age, "years old!")
}
say_hi(name = "Jillian", age = 22)

```


### Example 3: animal ages with if else 
```{r}
convert_animal_age <- function(animal_species, human_age){
  if (animal_species == "dog"){
    print(human_age * 7)
  } else if (animal_species == "goat"){ 
    print(human_age * 4.7)
  } else {
    print("This animal input is not dog or goat!")
  }
}
  
convert_animal_age("goat", 47)

convert_animal_age("monkey", 40) # returns nothing... for now! 
## future jillian speaking: now it returns my very own error message 
```

### Example 4: Teddy's favorite foods, tribble() function, and making a function using data frame values!

```{r}
dogs_favorite_foods <- tribble(
  ~dog_name, ~food, 
  "Khora", "everything",
  "Teddy", "salmon",
  "Waffle", "pancakes",
  "Banjo", "chicken"
)

dog_menu <- function(name) {
  my_sub <- filter(dogs_favorite_foods, dog_name == name)
  print(paste0("My name is ", my_sub$dog_name, " and I love ", my_sub$food, "!"))
}

dog_menu(name = "Waffle")

```

#### Example 5: Functions meet for loops (basically creating the apply function) -- print means of every column with a nice statement 

```{r}
means_over_columns <- function(df) {
  for(i in seq_along(df)) {
    col_mean <- mean(df[[i]], na.rm = TRUE)
    column_name <- colnames(df[i])
    print(paste("The mean value of", column_name, "is", col_mean))
  }
}

# fun example with an actual fully numeric data frame ! 
means_over_columns(df = mtcars)
means_over_columns(df = faithful)
```

